ASTR480 Final Report:\\ \textit{Asteroids in TESS}

### 2024-09-26 FinalReport/FinalReportDraft.tex
ios chat work 26/9

Using the \texttt{KDTree} algorithm \citep{Maneewongvatana1999} as implemented in \texttt{SciPy} \citep{2020SciPy-NMeth}, the RA and Dec coordinates of the interpolated points and the detections can be compared and matched together.
Filtering this \texttt{KDTree} output by restricting the time between spatially coincident matches to less than \qty{0.1}{\day} stops any accidental matches in position from non-asteroid detections.
A cut is then made by smallest distance, in coordinate space, and any distance must be smaller than \qty{0.01}{\degree} (\qty{36}{\arcsec}), which is within 2 TESS pixels.
Then, if multiple detection match to the same interpolated point, the detection with the smallest distance will be taken.
An example of the matches found using this \texttt{KDTree} is shown in \autoref{Fig:RADecMatch}.

@@

%light curves; detected VS forced interpolated 
There are two sets of points to take light curves from.
The matches from the detections, which already have a flux calculated, and the interpolated points themselves, which are more numerous but require forcing the photometry.

There were some challenges getting the flux of the interpolated points, even when \texttt{TESSELLATE} has already reduced all the FFIs of interest.
The flux is summed in  $3\times3$ box centred on the integer $xy$ position returned from changing the world coordinates in RA and Dec to pixel coordinates using \texttt{Astropy}'s \texttt{SkyCoord} subpackage. %? cite

The integer coordinates of the asteroids seemed to be leading the object in any given frame of TESS. %?FIG?
This was decreasing the total flux of the asteroid, as the box used for sumation of the fluxes was hitting mostly sky.
This also caused a ``sawtoothing'' effect, where the lighturve would jump from a minimum to a maximum when the integer coordinate snaped from being a pixel out to back on the asteroid.
This both of these issues can be fixed by calculation of the centre of mass (COM) flux, the average brightness in $xy$ space on a $8\times8$ cut of the image centred on the interpolated position.
I didn't take the highest pixel value in this box because there could be a rouge star, this would drag the COM around, but not by as much as snapping to that highest point.
A \texttt{Photutils} %TODO cite 
aperture photometry call to this COM with a radis of \qty{1.5}{\px} was enough to get fluxes of the same size, but with a far more smooth transition from maximum to minimum, as expected from rotating asteroids.

Not every interpolated point gets a match, due to a variety of reasons, which culminate in there being not an extreme enough of a difference between frames.
A comparison between the two light curves is interesting.

\autoref{Fig:DifFlux} shows these two light curves for a chosen asteroid, ``Ruff''.
This was picked as it has a high number of points matched to detections, but of course still more interpolated points total.
There seems to be a systematic offset between the fluxes, with the interpolated points having consistently lower flux, (here the means differ by 58 counts)
Adjusting aperture to sum the fluxes over to the ``centre of mass'' of the asteroid in each frame does not alleviate this problem.

%TODO data cleaning explaination
Following \citet{McNeill2023}, any fluxes more than $3\sigma$ from the mean value are sigma-clipped from the lightcurves.

@@

where the flux, $F$, is given in terms of time, $t$, the calculated frequency, f, and the model parameter $\vec{\theta} = [\theta_0, \theta_1,\theta_2]$.
These parameters are useful in reconstructing the sinusoid without rerunning the periodogram.

There will be differences in the period between the matched points and the interpolated points, not just because of the difference in average flux, but also from the larger number of points. This difference needs to be carefully thought through to understand what is the more likely period.

@@

%Run on server
The \texttt{TESSELLATE} pipeline has been running on the OzSTAR supercomputing facilities.
After I am confident that all the parts of the asteroid detection and subsequent light curve analysis works as required, the same code can be refactored to work on OzSTAR and a large-scale analysis of all the processed TESS sectors can be run.
Only after this has completed can the asteroid population statistics can be computed.
I will be looking for completeness of detections of asteroids, as well as accuracy of periods and amplitude variation.

The completeness of the asteroid population in TESS as a function of their apparent magnitude can be determined.
The apparent magnitude is one of the properties returned from the queries, and is in the V band.
By taking a limit on the if an asteroid is detected \dots %TODO what is the limit
Breaking the average V mag of these objects into \qty{0.5}{\mag} bins for both the sample returned from a query and the sample detected, a ratio of completeness can be calculated per bin.
(Note: If there are no asteroids in a bin, none have a chance of being recovered, so the all did get recovered. I.e. $\frac{0}{0}=1$)

@@

>>>>

Using the \texttt{KDTree} algorithm \citep{Maneewongvatana1999} as implemented in \texttt{SciPy} \citep{2020SciPy-NMeth}, the RA and Dec coordinates of the interpolated points and the detections can be compared and matched together.
Filtering this \texttt{KDTree} output by restricting the time between spatially coincident matches to less than \qty{0.1}{\day} stops any accidental matches in position from non-asteroid detections.
A cut is then made by smallest distance, in coordinate space, and any distance must be smaller than \qty{0.01}{\degree} (\qty{36}{\arcsec}), which is within 2 TESS pixels.
Then, if multiple detection match to the same interpolated point, the detection with the smallest distance will be taken.
An example of the matches found using this \texttt{KDTree} is shown in \autoref{Fig:RADecMatch}.
It is clear that \texttt{TESSELLATE} is not detecting every frame that an asteroid is in.
Detctions in \texttt{TESSELLATE} are done on a per pixel basis, and asteroids only cross each pixel for a short number of frames.
Not every interpolated point gets a match, due to a variety of reasons, which culminate in there being not an extreme enough of a difference between frames at that pixel location.
Instead there are bunches of freqntly matched points, and then patches where there are few matches detected at all.

%TODO fig for agregate of all asteroids in a sector, to go in results

@@

%light curves; detected VS forced interpolated 
There are two sets of points to take light curves from.
The matches from the detections, which already have a flux calculated, and the interpolated points themselves, which are more numerous but require forcing the photometry.
There were some challenges getting the flux of the interpolated points, even when \texttt{TESSELLATE} has already reduced all the FFIs of interest.

Using the \texttt{SkyCoord} module of \texttt{Astropy}, the world coordinates, in RA and Dec, can be transformed into pixel coordinates, $x$ and $y$, using a world coordinates system (WCS) from TESSELLATE.
\texttt{Photutils} \citep{Bradley2024} was then used to do aperture photometry with a \qty{1.5}{\px} radius to account for the spread in the TESS PRF. %?cite PRF

The integer coordinates of the asteroids seemed to be leading the object in any given frame of TESS. %?FIG?
This was decreasing the total flux of the asteroid, as the box used for sumation of the fluxes was hitting mostly sky.
This also caused a ``sawtoothing'' effect, where the lighturve would jump from a minimum to a maximum when the integer coordinate snaped from being a pixel out to back on the asteroid.
This both of these issues can be fixed by calculation of the centre of mass (COM) flux, the average brightness in $xy$ space on a $5\times5$ cut of the image centred on the interpolated position.
I didn't take the highest pixel value in this box because there could be a rouge star in the field that it snaps to, instead of the asteroid.
A \texttt{Photutils} aperture photometry call to this COM with a radius of \qty{1.5}{\px} was enough to get fluxes of the same order, but with a far more smooth transition from maximum to minimum, as expected from rotating asteroids.

A comparison between the two light curves is interesting, as one might expect only the brightest parts of the lightcurve will be matched.
\autoref{Fig:DifFlux} shows these two light curves for a chosen asteroid, ``Ruff''.
This was picked as it has a high number of points matched to detections, but of course still more interpolated points total.
There seems to be a systematic offset between the fluxes, with the interpolated points having consistently lower flux, (here the means differ by 58 counts)  %!check

%TODO data cleaning explaination
Following \citet{McNeill2023}, any fluxes more than $3\sigma$ from the mean value are sigma-clipped from the forced photometry lightcurves.
This does introduce gaps into an otherwise regularly sampled lightcurve, but the contamination from background stars the object is passing in front of is too great.

@@

where the flux, $F$, is given in terms of time, $t$, the calculated frequency, f, and the model parameter $\vec{\theta} = [\theta_0, \theta_1,\theta_2]$.
These parameters are useful in reconstructing the sinusoid without rerunning the periodogram.
The method used for the periodogram fits is the new \texttt{nifty-ls} package \citep{Garrison2024} that has can be implemented on its own, or by interfacing with \texttt{Astropy}.
I use the later implementation as the speed up from the nonuniform fast Fourier transform is helpful on the large dataset from \texttt{TESSELLATE}, but the astropy interface is how I had everything set up to run on just one cut.

I an using a frequency window of for the periodogram between the the inverse of twice the time the asteroid spends in the sector and the Nyquist frequency (1/TESS cadance).

%???
There will be differences in the period between the matched points and the interpolated points, not just because of the difference in average flux, but also from the larger number of points. This difference needs to be carefully thought through to understand what is the more likely period.

@@

%Run on server
The \texttt{TESSELLATE} pipeline has been running on the OzSTAR supercomputing facilities.
After I was confident that all the parts of the asteroid detection and subsequent light curve analysis worked as required on one cut of one camera of one ccd of one TESS sector, the same code was be refactored slightly to work on OzSTAR and a large-scale analysis of all the processed TESS sectors can be run.

I will be looking for completeness of detections of asteroids, as well as accuracy of periods and amplitude variation by comparison to the LCDB.

The completeness of the asteroid population in TESS as a function of their apparent magnitude can be determined.
The apparent magnitude is one of the properties returned from the queries, and is in the V band.
By taking a limit on the if an asteroid is detected \dots %TODO what is the limit
Breaking the average V mag of these objects into \qty{0.5}{\mag} bins for both the sample returned from a query and the sample detected, a ratio of completeness can be calculated per bin.
(Note: If there are no asteroids in a bin, none have a chance of being recovered, so the all did get recovered. I.e. $\frac{0}{0}=1$, this is only an issue at the brightest magnitude where the fewest asteroids exist.)

@@

This research made use of Photutils, an Astropy package for detection and photometry of astronomical sources \citep{Bradley2024}.
